---
title: "Applying geographical filters to MOTUs lists using geotag"
author: "Nicolas Le Guillarme"
output: rmarkdown::html_vignette
runtime: shiny
vignette: >
  %\VignetteIndexEntry{taxonerd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Learning objectives

This vignette is designed to:
1. show how to install the geotag package for R
2. show how to initialize the taxonomic reference
3. show how to apply a geographical filter on a list of MOTUs based on their GBIF occurrences

# Install geotag

geotag is a R package that provides geographic filters to clean up lists of MOTUs based on prior knowledge of their geographic distribution.

You can always install the latest release of geotag from GitHub with the following command:

```{r eval=FALSE}
install.packages("https://github.com/nleguillarme/geotag/releases/latest/download/geotag.tar.gz", repos=NULL)
```

You can also check for available releases on the project's GitHub page : https://github.com/nleguillarme/geotag/releases

Now that geotag is correctly installed, you are ready to go.

# Initialize the taxonomic reference

geotag uses the taxonbridge package (https://github.com/MoultDB/taxonbridge) to merge the GBIF backbone and NCBI taxonomies into a single taxonomic reference. This allows geotag to handle both taxonomies seamlessly.

The first time you use geotag, you have to build this taxonomic reference using the following command:

```{r eval=FALSE}
prepare.taxonomy()
```

This will create a file "taxonomy.rds" in the working directory.

# Apply the GBIF filter to a list of MOTUs

Before using one of the filters provided by geotag, you have to load the taxonomic reference:

```{r}
taxonomy = readRDS("taxonomy.rds")
```

Then load your data:

```{r}
df = read.csv("data/sample.csv")
```

The data frame contains a column called "taxa_to_filter", which contains the scientific names of the MOTUs in the NCBI taxonomy.

```{r}
df$taxa_to_filter
```

The next step is to define the geographical area in which you wish to search for GBIF hits. This geographical area can be a country, specified by its 2-letter country code (ISO-3166-1), or an arbitrary polygon in Well Known Text (WKT) format. See the documentation of rgbif for more details.

```{r}
# The geographical area can be a polygon in WKT format...
geometry = "POLYGON((-75.61464 62.21534,-78.7656 60.5815,-77.61418 56.70173,-79.45098 54.67159,-87.65469 55.59753,-93.43596 57.88686,-95.31212 60.04932,-129.56689 53.81444,-123.45217 45.93347,-123.91768 40.93663,-118.11493 34.16381,-111.4253 31.22765,-100.18644 27.59835,-93.76801 30.41739,-83.49371 29.99948,-76.59172 36.77585,-70.69512 42.58472,-65.44863 46.67466,-65.75653 49.56489,-57.72713 47.71301,-52.93458 47.64121,-57.82193 54.45649,-64.46424 60.35065,-67.60679 59.05623,-75.61464 62.21534))"
# ... or a 2-letter country code (ISO-3166-1)
country = "fr"
```

Finally, run the GBIF filter on your data. This requires specifying the name of the column (col.name) containing the MOTUs information, whether this information is a taxonomic identifier or a scientific name (taxid = TRUE/FALSE), and what is the reference taxonomy (ref.taxo = "ncbi"/"gbif"). You can also specify the minimum number of hits (min.occ) for a MOTU to be considered present in the geometry/country.

```{r}
filtered = gbif.filter(df, col.name="taxa_to_filter", geometry=geometry, country=NULL, min.occ = 1, taxid=FALSE, ref.taxo="ncbi", taxonomy=taxonomy)
```

The GBIF filter adds a column "gbif_filter" to the input data frame, containing TRUE if the MOTU is present in the geometry/country, FALSE if not, and NA if the MOTU's name/taxid could not be matched to a taxon in the GBIF backbone.

```{r}
subset(filtered, select = c("taxa_to_filter", "gbif_filter"))
```
